Joint estimation of phylogenies and ancestral states is not consistent

### 2017-12-18 appendix.tex
remove etas, clarify lower bound, identifiability

Let
$$
%EM It seems a little strange that eta appears here as an argument when it didn't appear in eq:site_pattern_likelihood. In fact it's not really an argument, right? Everything is determined by t and tau it seems to me.
%das: We added it to clarify that the likelihood now depends on \xi, which is an element of \eta. It seems dirty to ignore it. At the risk of adding more notation, we could add \hat{H} to the original likelihood? Or just axe the etas...
%EM I think I must be missing something:
%EM The likelihood doesn't belong on the specific choice of \xi. Besides, \xi already appears in eq:site_pattern_likelihood.
%EM Eta is a deterministic function of t and tau, so isn't having it redundant?
%das: The likelihood doesn't depend on the specific choice of \xi, but it does depend on \eta. I don't know, I guess redundancy isn't always the worst thing? I just see it as a tradeoff between how explicit we want to be and how clean. Like, in the first line of eq:site_pattern_likelihood we could have as an argument \hat{H} instead of suppressing it, even though it's totally determined by the inputs as-is. Anyway, that was my justification for including \eta. I'm fine with removing it though if you think it's weird.
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\end{equation}

@@

%EM I think this following statement could use a little more justification. If we write out a symmetric function f in two variables and take the derivative of f(x+delta, x) = f(x, x+delta) with respect to delta and set it to zero we can see that this must be a stationary point of f. Do we need to say more to ensure that this is a maximum?
%das: I wasn't even thinking derivatives. Let f(x, y) be symmetric so that f(x, y) = f(y, x). If f is maximized at some (x_0, y_0) then f(x_0, y_0) > f(x, y) for all x, y. But then by symmetry f(y_0, x_0) is also a maximum. Now *if* a unique maximum exists, x_0 = y_0 should necessarily be this maximum by the properties defined above, otherwise, f(x_0, y_0) would not be equal to f(y_0, x_0). I think there are probably counterexamples where there is not a unique maximum where this statement isn't necessarily true (trivially, because you'd just say x_0 = y_1 if y_1 is also a maximum and voila) but I don't think this discounts the fact that x_0 = y_0 will still be *a* maximum in this case. In that case, it's reasonable to assume so, in my opinion, especially for the topological inconsistency (we don't use the branch length values), but even for the branch length inconsistency because the case of no unique solution is just as bad if not worse than inconsistency.
%das: Should some or all of that be in the text? Can you see anything wrong with it?
%EM I don't see why this works. In fact, imagine that x=y is a ridge of minima, and that we have maxima at (1,2) and at (2,1). My argument shows that we must have a stationary point at x=y, but that doesn't mean that it has to be a maximum!
%das: Wouldn't the assumption of there existing a unique maximum preclude your case? My argument is that if there is a unique maximum, it necessarily has to have x=y. If there is not a unique maximum then the ML problem is ill-posed. I now don't see why we have to have a unique maximum other than to say we hope the problem is well-posed, but it may not be. Do you have any thoughts on where to take this?
Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, $x_1=x_2$ at the maximum.
The analogous statement holds for $y_1$ and $y_2$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

In the Felsenstein case, there are quite a few optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
We can bound these below replacing $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$, as this will result in a lower bound on the maximum, so we proceed with that first.
Then, for site split $\{1,2\}$ we have

@@

always, the maximum can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.

>>>>

Let
$$
%EM It seems a little strange that eta appears here as an argument when it didn't appear in eq:site_pattern_likelihood. In fact it's not really an argument, right? Everything is determined by t and tau it seems to me.
%das: We added it to clarify that the likelihood now depends on \xi, which is an element of \eta. It seems dirty to ignore it. At the risk of adding more notation, we could add \hat{H} to the original likelihood? Or just axe the etas...
%EM I think I must be missing something:
%EM The likelihood doesn't belong on the specific choice of \xi. Besides, \xi already appears in eq:site_pattern_likelihood.
%EM Eta is a deterministic function of t and tau, so isn't having it redundant?
%das: The likelihood doesn't depend on the specific choice of \xi, but it does depend on \eta. I don't know, I guess redundancy isn't always the worst thing? I just see it as a tradeoff between how explicit we want to be and how clean. Like, in the first line of eq:site_pattern_likelihood we could have as an argument \hat{H} instead of suppressing it, even though it's totally determined by the inputs as-is. Anyway, that was my justification for including \eta. I'm fine with removing it though if you think it's weird.
%das: I changed all this.
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\end{equation}

@@

%EM I think this following statement could use a little more justification. If we write out a symmetric function f in two variables and take the derivative of f(x+delta, x) = f(x, x+delta) with respect to delta and set it to zero we can see that this must be a stationary point of f. Do we need to say more to ensure that this is a maximum?
%das: I wasn't even thinking derivatives. Let f(x, y) be symmetric so that f(x, y) = f(y, x). If f is maximized at some (x_0, y_0) then f(x_0, y_0) > f(x, y) for all x, y. But then by symmetry f(y_0, x_0) is also a maximum. Now *if* a unique maximum exists, x_0 = y_0 should necessarily be this maximum by the properties defined above, otherwise, f(x_0, y_0) would not be equal to f(y_0, x_0). I think there are probably counterexamples where there is not a unique maximum where this statement isn't necessarily true (trivially, because you'd just say x_0 = y_1 if y_1 is also a maximum and voila) but I don't think this discounts the fact that x_0 = y_0 will still be *a* maximum in this case. In that case, it's reasonable to assume so, in my opinion, especially for the topological inconsistency (we don't use the branch length values), but even for the branch length inconsistency because the case of no unique solution is just as bad if not worse than inconsistency.
%das: Should some or all of that be in the text? Can you see anything wrong with it?
%EM I don't see why this works. In fact, imagine that x=y is a ridge of minima, and that we have maxima at (1,2) and at (2,1). My argument shows that we must have a stationary point at x=y, but that doesn't mean that it has to be a maximum!
%das: Wouldn't the assumption of there existing a unique maximum preclude your case? My argument is that if there is a unique maximum, it necessarily has to have x=y. If there is not a unique maximum then the ML problem is ill-posed. I now don't see why we have to have a unique maximum other than to say we hope the problem is well-posed, but it may not be. Do you have any thoughts on where to take this?
%das: So there are references for identifiability, but I'm a little nervous using them because they're for the usual marginal ML problem. How cagey does it sound to say (as I do below) "here are references for identifiability in the usual case, so either joint inf will be not identifiable (bad) or it (likely?) will be and we can use this (which turns out to be inconsistent later)?" Can you think of a better approach?
%das: some other references I found for identifiability are Rogers 1997, RoyChoudhury 2014...
The maximum likelihood procedure in phylogenetics results in identifiable parameter estimation under mild conditions \cite{Chang1996}.
The joint inference procedure may not necessarily result in the same identifiability properties, but we claim the following.
If joint inference obtains unique parameter estimates, then, since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, $x_1=x_2$ at the maximum, with the analogous statement holding for $y_1$ and $y_2$.
The case in which we do not obtain unique parameter estimates results in an ill-posed likelihood problem, and thus by design an inconsistency since multiple maxima exist that will not equal the true, generating parameters.
Given this, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property of symmetry of function inputs, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

In the Felsenstein case, there are many more optimal internal states depending on branch lengths than in the Farris case (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).
We first bound the entire likelihood below and then proceed with joint inference.
First, we obtain a lower bound for the likelihood by replacing $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$; in this case, we resolve many of the ambiguous likelihood terms.
For example, for site split $\{1,2\}$, after we replace the $1+w$ term for $\ancestralSplitPartition_j(\tau, t)=\{\emptyset, \{1,2\}\}$ with $1-w$, we have

@@

%das: Is this better?
always, the maximum for this site split can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
